The Concurrent Calculi Formalisation Benchmark

### 2024-02-20 short-paper/short-paper.tex
Revise section on linearity

\todo{Add an informal description of why we care about linearity here}

For our challenge on linear reasoning we have chosen a type safety
theorem for a system using session types, combining as usual subject reduction with absence of errors.  This allows us to introduce
linear reasoning with few definitions.  With this focus, we forgo
channel delegation: input and output receives/sends only values, which
do not include channels, hence the topology of the communication
network described by a process cannot change.  We do not allow
recursion or replication in the syntax, hence no infinite behaviours
can be expressed. In particular, we only model linear (as opposed to
shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names, highlighting their
duality within the type system. The latter enforces
 two invariants, namely that no endpoint is used simultaneously by parallel processes and
 that the two endpoints of the same session have dual types.
The operational semantics is formulated as a standard reduction relation to factor the subject reduction proof,  encompassing type
preservation for structural congruence. W.r.t.\ type safety, we have
chosen a simple notion of what an error is: well-typed processes are a
well-formed in the sense that they do not use endpoints in a non-dual
way.  While more sophisticated notions of well-formedness that
prevent \eg deadlocks are interesting, the proofs of those properties
would distract from the core linear aspects of the challenge.

In mechanised meta-theory, addressing linearity means choosing an
appropriate representation of the linear context.  The crux of the
matter is \emph{context splitting}, as showcased by the typing rule
for parallel composition.  While a linear context is perhaps best seen
as a multi-set, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal
to the mathematics of the problem under study, in our case proving
type safety for session types.  In this space, several designs are
possible: one can label occurrences of resources to constrain their
usage (\eg~\cite{CicconeP20}); some authors instead impose a multiset
structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can
be implemented as finite maps (as in \cite{Castro2020}), whose operations
are sensitive to a linear discipline. In all these cases, the effort
required to develop the infrastructure is significant.

A different approach is to bypass the problem of context splitting by adopting familiar ideas from algorithmic linear type checking: this is known as \emph{typing with leftovers}, as exemplified in~\cite{DBLP:conf/forte/ZalakainD21}.
Whatever the choice, list-based encodings can be refined to be intrinsically-typed, if the proof assistant supports dependent types (see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).

A completely different approach is to adopt a \emph{sub-structural} meta-logical framework: here the framework itself handles resource distribution, and users need only map their linear operations to the ones offered by the framework.
The linear function space will then handle all context operations implicitly, including splitting and substitution.
One such framework is \emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session types in~\cite{Bock2016}).
Unfortunately, \emph{Celf} does not yet fully support the verification of meta-theoretic properties.
A compromise is the so called \emph{two-level} approach, where one encodes a sub-structural specification logic in a mainstream proof assistant and then uses that logic to state and prove linear properties (for a recent example, see~\cite{Felty:MSCS21}).

>>>>

Linear typing systems allow for tracking the use of resources in a program. In
the case of typed process calculi (in particular, session types), linearity is
widely used for tracking if and how a channel is used to send or receive values.
This leads to substructural typing~\cite[Chapter 1]{Pierce2004-oq} with
resulting mechanisation difficulties --- \eg deciding how to \emph{split the
typing context} in a parallel composition.

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen to only model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
% The latter enforces two invariants, namely that no endpoint is
% used simultaneously by parallel processes and that the two endpoints of the same
% session have dual types.
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

We designed this challenge to focus on linear reasoning while minimising
definitions and other concerns.  Therefore, we forgo channel delegation:
send/receive operations only support values which do not include channels, hence
the topology of the communication network described by a process cannot change.
We do not allow recursion or replication, hence no infinite behaviours can be
expressed. We also forgo more sophisticated notions of error-freedom (\eg
deadlock freedom) because proving them would distract from the core linear
aspects of the challenge.

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multi-set, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}); some
authors instead impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%
A different approach is to bypass the problem of context splitting by adopting familiar ideas from algorithmic linear type checking: this is known as \emph{typing with leftovers}, as shown in~\cite{DBLP:conf/forte/ZalakainD21}.
Whatever the choice, list-based encodings can be refined to be intrinsically-typed, if the proof assistant supports dependent types (see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%
A completely different approach is to adopt a \emph{substructural} meta-logical framework: the framework itself handles resource distribution, and users need only map their linear operations to the ones offered by the framework.
The linear function space will then handle all context operations implicitly, including splitting and substitution.
One such framework is \emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session types in~\cite{Bock2016});
unfortunately, \emph{Celf} does not yet fully support the verification of meta-theoretic properties.
A compromise is the so called \emph{two-level} approach, \ie encoding a substructural specification logic in a mainstream proof assistant and then using that logic to state and prove linear properties (for a recent example, see~\cite{Felty:MSCS21}).
